Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2022-08-02 paper/ch_potentials.tex
polishing

\begin{abstract}
We study exact, efficient and practical algorithms for route planning applications in large road networks.
Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the specific application.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
A* is a classical approach to accelerate Dijkstra's algorithm.
A* can support many extended scenarios without much additional implementation complexity.
However, A*'s performance depends on the availability of a good heuristic that estimates distances.
Computing tight distance estimates is a challenge on its own.
On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
They achieve speed and exactness but sacrifice A*'s flexibility.
Extending them to certain practical applications can be hard.
In this paper, we present Lazy RPHAST, an algorithm to efficiently compute distances from many vertices towards a single target and use it as an A* heuristic.
As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
Our approach allows decoupling the supported applications from the hierarchical speedup technique.
Additionally, we describe A* optimizations to accelerate the processing of low degree vertices, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
An extensive experimental study confirms the practical applicability of our approach for many applications.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
The past decade has seen a plethora of research works on route planning in large road networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Vertices represent intersections.
Roads are modeled using edges.
Edges are weighted by their traversal times.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
Unfortunately, on continental-sized networks, it is too slow for many applications.
Therefore, speedup techniques have been developed.
These techniques employ an offline \emph{preprocessing} phase where auxiliary data is precomputed.
This auxiliary data is then utilized to accelerate shortest path computations in the online \emph{query} phase.
With this approach, speedups of more than three orders of magnitude have been achieved.
This allows interactive query times of milliseconds or less on continental-sized road networks.

However, for many real world applications, this basic model is too simplistic.
For realistic routing, many additional features need to be considered.
This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions.
Some applications may have additional application-specific requirements.
Extending Dijkstra's algorithm to support these features is usually easy.
In contrast, extending speedup techniques is vastly more complex.
The results on extending specific speedup techniques to specific extended problem fill many research papers~\cite{dgpw-crprn-13} and sometimes entire dissertations~\cite{b-tdrpc-14}. % TODO cite some papers
We will present a brief overview in Section~\ref{sec:related_work}.
While techniques achieving fast query times have been successfully developed for many extended scenarios, we observe two problems:
The resulting techniques are complex to implement and hard to further extend and adjust.
For example, while there are efficient speedup techniques for each of the above mentioned features, we are not aware of any work supporting the combination of these features.
We believe that for \emph{practical} algorithms, a different trade-off is necessary.
Therefore, in this work, we prioritize a simple and extensible approach over the fastest possible query times.

@@

In the past two decades, a considerable amount of research effort has been put into accelerating shortest path computations in transportation networks.
For a comprehensive overview, we refer to~\cite{bdgmpsww-rptn-16}.
Here, we first highlight a few key results for the classical shortest path problem on road networks specifically relevant to our work.
In the second part of this section, we will discuss adaptions of these techniques to extended problem models.

\paragraph{Speedup Techniques for Shortest Paths in Road Networks}
A*~\cite{hnr-afbhd-68} is a well-known approach to accelerate shortest path queries by directing the search towards the target through heuristic distance estimates.
It has been successfully employed in many problems beyond route planning in transportation networks~\cite{DBLP:conf/socs/StrasserHB14,DBLP:conf/ijcai/BonoGHS19,DBLP:conf/ijcai/0002UJAKK18}.
It is one of the algorithms fundamental to our work.
The performance of A* crucially depends on the quality of the heuristic estimates.
A natural but rather ineffective heuristic can be obtained through euclidian distances.
On road networks, it performs worse than Dijkstra's algorithm~\cite{gh-cspas-05}.
A much more accurate heuristic can be obtained with ALT~\cite{gh-cspas-05,gw-cppsp-05}.
ALT was one of the early speedup technique for routing in road networks.
It uses precomputed distances to certain landmark vertices combined with the triangle inequality to compute distance estimates.
On road networks, this achieves speedups of around two orders of magnitude over Dijkstra's algorithm.
However, the speedups obtained by ALT are still limited by the quality of the distance estimates.
To achieve faster queries, Core-ALT has been developed~\cite{bdsssw-chgds-10}.
Here, parts of the network are \emph{contracted} during the preprocessing, leaving a much smaller remaining core graph to run the goal directed search on.
This yields speedups of about three orders of magnitude compared to Dijkstra's algorithm.
Since then, there has been little development on A*-based techniques for routing in road networks.
This is because hierarchical techniques have proven much more effective.

Hierarchical speedup techniques utilize the inherent hierarchy in road networks, usually in combination with the \emph{contraction} of less important vertices.
One popular example is Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network at query time.
The ``importance'' of vertices is determined heuristically during preprocessing.
This results in queries taking less than a tenth of a millisecond which is a speedup of four or more orders of magnitude over Dijkstra's algorithm.
The preprocessing takes only a few minutes.
CH has been used successfully in many real world applications and quite a few open-source implementations exist\footnote{
\url{https://gist.github.com/PayasR/bc46af938195a827e42006c3f5544e4a}.
While this list is certainly not exhaustive and possibly not representative for what algorithms are used in cooperate contexts, it still paints a fairly clear picture on which techniques have reached a certain popularity in practice.
}.
CH is another fundamental algorithm for our work.
More specifically, we extend on PHAST~\cite{dgnw-phast-13} and RPHAST~\cite{delling_et_al:OASIcs:2011:3266}.
These algorithms reuse the CH preprocessing but modify the query phase to efficiently support \emph{one-to-all} queries from one source vertex to \emph{all} other vertices (PHAST), or \emph{one-to-many} queries to a subset of vertices known in advance (RPHAST).
%
Another popular hierarchical speedup technique also used in practice~\cite{bingblog} is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
It is based on a multi-level partition of the road network computed during preprocessing.
Queries are slightly slower than CH queries but still three to four orders of magnitude faster than Dijkstra's algorithm.
Hierarchical speedup techniques have been combined with goal-directed search and A* in~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, these works also primarily focused on obtaining faster queries rather than simple and flexible approaches.

The speedup technique known to currently achieve the fastest queries is Hub Labels (HL)~\cite{TODO}.
With HL, query times in less than a microseconds are possible.
This is only a couple of times the time of an uncached memory access.
This comes at the cost of an expensive preprocessing phase and a tremendous memory footprint, more than an order of magnitude larger than the input graph.
Despite the extremely fast query times, it appears that the trade-off offered by HL is, in practice, often not very attractive.
Simpler techniques like CH already offer queries fast enough to completely disappear behind other parts of practical applications (network latency, etc.).
Also, we are not aware of any works extending HL to extended problem models.

% extended application
\paragraph{Speedup Technique Adaptions for Extended Problems}
One of the earliest approaches to adapt speedup techniques to extended problems was based on A*.
In~\cite{dw-lbrdg-07}, ALT is used for routing on dynamic and time-dependent graphs.
This approach is conceptually similar to ours.
However, because the heuristic is ALT-based, the accuracy of the estimates is limited and the resulting query performance not competitive.
Further, the evaluation was performed only with synthetic traffic data.
With production-grade real-world traffic data, the problem becomes significantly harder~\cite{swz-sfert-21}.
Similar to the development for the classical problem, the approach was also combined with contraction~\cite{ndls-bastd-12,dn-crdtd-12}.
Even though this improves query running times, they obtained speedups are not competitive with purely hierarchical techniques.

CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of MLD~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
For this, a second, faster, preprocessing phase is introduced which takes at most a few seconds.
This phase is called the \emph{customization}.
It can be run regularly to update weights.
This enables the integration of live traffic and user preferences.
CRP was designed to support turn costs without any additional modifications.
Queries in CRP take at most a few milliseconds.
CRP is one of the few examples of a speedup technique designed for flexibility rather than maximum query performance.
However, its flexibility is also limited.
Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
Achieving reasonable memory consumption and query times was only possible by giving up exactness.
Further, TD-CRP can only compute approximate shortest distances (rather than paths).

There is also lot of work that extends CH to more complex problem models.
In~\cite{gv-errnt-11}, turn information is integrated into CH.
The proposed approach achieves fast queries, but preprocessing becomes an order of magnitude slower compared to classical CH.
In~\cite{dsw-cch-15}, CH is extended to Customizable Contraction Hierarchies (CCH).
CCH also has a second preprocessing (customization) phase where weights can be altered.
This allows to support user preferences and live traffic with queries an order of magnitude faster than CRP.
Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
Even with all improvements proposed in~\cite{bwzz-cchtc-20}, turn integration still causes a slowdown of at least factor three to the customization phase.
A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
Several papers~\cite{bdsv-tdch-09,bgns-tdcha-10,klsv-dtdch-10,bgsv-mtdtt-13} and an entire dissertation~\cite{b-tdrpc-14} have been published on the subject.
Different variants with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
Recently, a new approach has been published~\cite{swz-sfert-21} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity. % TODO argue what is complicated to implement?

@@

Clearly, there is a plethora of efficient speedup techniques for routing in road networks.
Further, for many extended problems there are dedicated research results on extensions of some techniques.
While many problems can be solved very efficiently, we still believe that this situation is not satisfactory in practice.
For practical applications, a simple, unified, and extensible approach is often more important than the fastest query performance.
Therefore, in this work, we revisit the A* algorithm and build a flexible, unified framework for a multitude of routing problems.
The core of our approach is a new CH-based A* heuristic.
It allows for much tighter estimates than previous A* heuristics and thus significantly faster queries.
While the query running times of our approach are for the most part not competitive with approaches significantly tailored to specific problems, they are still sufficient for practical applications.
The remainder of this work is organized as follows:

We discuss notation and fundamental algorithms in Section~\ref{sec:preliminaries}.
In Section~\ref{sec:lazy-rphast}, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem.
Section~\ref{sec:astar_opts} contains several optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
Our main contribution, the CH-Potentials framework, is presented in Section~\ref{sec:framework}.
There, we first show how to use Lazy RPHAST to build a fast and tight heuristic for A* in road networks.
Based on this heuristic, we provide a unified framework (Section~\ref{sec:formal_framework} and~\ref{sec:chpot}) for a variety of practical route planning problems (Section~\ref{sec:applications}).
CH-Potentials can be applied to all of these problems without any modifications to the preprocessing.
In Section~\ref{sec:experiments}, we provide an extensive experimental evaluation analyzing the performance characteristics of our algorithms.
It shows that CH-Potentials achieve decent running times when tight lower bounds are available at preprocessing time.
However, the strength of our approach is not in query running times but in its flexibility:
Problem extensions can be supported without complex algorithmic adjustments.
The price for each extension is a certain query slowdown depending on how tight the lower bounds used during preprocessing remain.

@@

The augmented graph has the property that for any two vertices $s$ and $t$, there always exists an \emph{up-down} $st$-path, i.e.\ a path  which first uses only edges from $\gchu$ and then only edges from $\gchd$, with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for an illustration.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Such an up-down path can be found by running Dijkstra's algorithm from $s$ on $\gchu$ and from $t$ on the reversed downward graph $\rgchd$ graph.
By construction, at least the highest ranked vertex on the up-down path must be in the intersection of the forward and backward search spaces.
Pseudocode for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
The set of vertices reachable in $\gchu$ and $\rgchd$ is called the \emph{CH search space} of a vertex.
The CH query performance strongly correlates with the size of this CH search space and the number of edges between the vertices in the search space.
Luckily, on road networks, the CH search space is small~\cite{gssv-erlrn-12,dgpw-crprn-13}.

@@

Some hierarchical techniques were extended to support a three-phase setup~\cite{dgpw-crprn-13}.
The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
The preprocessing phase is independent of the weights and only uses the graph topology.
Weights are introduced in the customization phase.
Paths are computed in the query phase.

Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) is a variant of Contraction Hierarchies that follows this three-phase setup.
The key idea is to use a vertex importance order that is independent of the weight function.
CCH use \emph{nested dissection} orders for this.
A small balanced vertex separator is computed and the separator vertices get the highest importance.
The remaining vertices are ordered by recursing on the independent cells obtained by removing the separator vertices.
This results in surprisingly good CH orders.
Regardless of the weight function, any shortest path between the cells has to use some of the separator vertices, therefore these vertices will always be ``important''.
Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19} in less than an hour even for continental-sized road networks.
Once the order was computed, the augmented graph $G^+$ is computed.
The standard CH preprocessing algorithms could be used for this but specialized, significantly faster CCH variants exist.
However, discussing them in detail is beyond the scope of this work.
See~\cite{dsw-cch-15,BuchholdSW19} for details.
A CCH augmented graph fulfills all necessary properties for CH query algorithms.
The CH query algorithms can therefore be applied without modifications.
However, the CCH preprocessing grants some useful additional properties.

During preprocessing, an \emph{elimination tree} can be constructed.
The root of the elimination tree is the most important vertex.
For every other vertex its parent is its least important upward neighbor.
In~\cite{bcrw-s-16}, it was proven that the set of ancestors of a vertex in the elimination tree is equal to its CH search space.
This makes it possible to explore the CCH search space with a more efficient shortest path algorithm:
For the forward search, start at the source and follow the elimination tree upward until the root is reached.
All outgoing edges of encountered vertices are relaxed.
The backward search works analogously but starts at the target.
Because this elimination tree-based algorithm does not require a queue, it is faster than a Dijkstra-based CCH query.
With this algorithm, CCH can consistently answer shortest path queries on continental sized road networks in fractions of a millisecond.
CH queries are still marginally faster for weight functions with a strong hierarchy such as travel times.
However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries turn out the faster variant.
Note that only nested dissection orderings admit an elimination tree of low height.
The elimination tree query can therefore not be used with a classical CH.

@@

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all vertices to one target vertex (or vice versa, the reverse case works analogously).
This is sometimes denoted as a \emph{all-to-one} (or \emph{one-to-all}) problem.
The preprocessing phase remains the same as for CH.
The query phase is split into two steps.
The first step is analogue to the CH query:
From $t$, all reachable vertices via reversed down-edges are explored, as shown in Algorithm~\ref{algo:ch-backward}.
For the second step, PHAST utilizes an assignment of vertices into \emph{levels}.
These levels correspond to the importance ordering but allow multiple vertices in the same level.
However, no edge must connect two vertices within the same level.
Such a level assignment can be obtained by assigning all vertices without lower ranked neighbors to the lowest level.
All other vertices are then iteratively assigned to the next level above the highest level of any downward neighbor.

The main work of the second step consists in iterating over all CH levels from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all levels are processed, the shortest distances from all vertices to $t$ were computed.
Pseudocode is provided in Algorithm~\ref{algo:phast}.
PHAST is faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures, better at utilizing data locality and, most importantly, can be parallelized very effectively.
% However, we will not consider parallelization in this paper.
See~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
% Using PHAST, we can also compute a tight A* heuristic.
% In the query phase, we first run PHAST to compute the distances from every vertex to $t$ with respect to $w$ and store the result in an array $H$.
% Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.

@@

We consider the combined running time of the target selection and each incremental query as the total running time to answer an incremental many-to-one query.
Since the target selection time is included, computing the distances to all vertices with Dijkstra or PHAST is too slow.
Also, since the source set is only provided incrementally, RPHAST in its basic form is not well suited to our problem.
Fortunately, we can do better.

>>>>

\begin{abstract}
We study exact, efficient and practical algorithms for route planning applications in large road networks.
On the one hand, such algorithms should be able to answer shortest path queries within milliseconds.
On the other hand, routing applications often require integrating the current traffic situation, planning ahead with predictions for future traffic, respecting forbidden turns, and many other features depending on the specific application.
Therefore, such algorithms must be flexible and able to support a variety of problem variants.
% There exists a lot of research focused on techniques to achieve the fastest possible query times.
% Further, there are a lot of results on extending certain techniques to specific scenarios and achieving fast queries in that specific extended problem.
% However, there is little research focused on designing extensible algorithms which can be adapted to new problem variants with minimal effort.
In this work, we revisit the A* algorithm to build a simple, extensible and unified algorithmic framework applicable to a variety of route planning problems. % without adjustments to the preprocessing.
A* has been previously used for routing in road networks.
However, its performance was not competitive because no sufficiently fast and tight distance estimation function was available.
We present a novel, efficient and accurate A* heuristic by utilizing Contraction Hierarchies (CH), another popular speedup technique.
The core of our heuristic is a new CH query algorithm called Lazy RPHAST which can efficiently compute shortest distances from many incrementally provided sources towards a common target.
Additionally, we describe A* optimizations to accelerate the processing of low-degree vertices, common in road networks, and present a new pruning criterion for symmetrical bidirectional A*.
An extensive experimental study confirms the practicality of our approach for many applications.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
The past two decades have seen a plethora of research works on route planning in large road networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Vertices represent intersections.
Roads are modeled using edges.
Edges are weighted by their traversal times.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
Unfortunately, on continental-sized networks, it is too slow for many applications.
Therefore, speedup techniques have been developed.
These techniques employ an offline \emph{preprocessing} phase where auxiliary data is precomputed.
This auxiliary data is then utilized to accelerate shortest path computations in the online \emph{query} phase.
With this approach, speedups of more than three orders of magnitude have been achieved.
This allows interactive query times of milliseconds or less on continental-sized road networks.

However, for many real world applications, this basic model is too simplistic.
For realistic routing, many additional features need to be considered.
This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions.
Some applications may have additional application-specific requirements.
Extending Dijkstra's algorithm to support these features is usually easy.
In contrast, extending speedup techniques is vastly more complex.
The results on extending specific speedup techniques to specific extended problem fill many research papers~\cite{gv-errnt-11,dgpw-crprn-13,dsw-cch-15,bwzz-cchtc-20,bgsv-mtdtt-13,swz-sfert-21} and sometimes entire dissertations~\cite{Delling2009_1000011046,b-tdrpc-14,Baum2018_1000082225}.
We will present a brief overview in Section~\ref{sec:related_work}.
While techniques achieving fast query times have been successfully developed for many extended scenarios, we observe two problems:
The resulting techniques are complex to implement and hard to further extend and adjust.
For example, while there are speedup techniques achieving fast queries for each of the above mentioned features, we are not aware of any work supporting the \emph{combination} of these features.
We believe that for practical algorithms, a different trade-off is necessary.
Therefore, in this work, we prioritize a simple and extensible approach over the fastest possible query times.

@@

A considerable amount of research effort has been put into accelerating shortest path computations in transportation networks.
For a comprehensive overview, we refer to~\cite{bdgmpsww-rptn-16}.
Here, we first highlight a few key results for the classical shortest path problem on road networks specifically relevant to our work.
In the second part of this section, we will discuss adaptions of these techniques to extended problem models.

\paragraph{Speedup Techniques for Shortest Paths in Road Networks}
A*~\cite{hnr-afbhd-68} is a well-known approach to accelerate shortest path queries by directing the search towards the target through heuristic distance estimates.
It has been successfully employed in many problems beyond route planning in transportation networks~\cite{DBLP:conf/socs/StrasserHB14,DBLP:conf/ijcai/BonoGHS19,DBLP:conf/ijcai/0002UJAKK18}.
It is one of the algorithms fundamental to our work.
The performance of A* crucially depends on the quality of the heuristic estimates.
For example, while geographic distances may seem like a natural choice for distance estimates in road networks, the resulting heuristic is rather ineffective.
In fact, it performs worse than Dijkstra's algorithm~\cite{gh-cspas-05}.
A much more accurate heuristic can be obtained with ALT~\cite{gh-cspas-05,gw-cppsp-05}.
ALT was one of the early speedup technique for routing in road networks.
It uses precomputed distances to certain landmark vertices combined with the triangle inequality to compute distance estimates.
On road networks, this achieves speedups of around two orders of magnitude over Dijkstra's algorithm.
However, the speedups obtained by ALT are still limited by the quality of the distance estimates.
To achieve faster queries, Core-ALT has been developed~\cite{bdsssw-chgds-10}.
Here, parts of the network are \emph{contracted} during the preprocessing, leaving a much smaller remaining core graph to run the goal directed search on.
This yields speedups of about three orders of magnitude compared to Dijkstra's algorithm.
Since then, there has been little development on A*-based techniques for routing in road networks.
This is because hierarchical techniques have proven much more effective.

Hierarchical speedup techniques utilize the inherent hierarchy in road networks, usually in combination with the \emph{contraction} of less important vertices.
One popular example is Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network at query time.
The ``importance'' of vertices is determined heuristically during preprocessing.
This results in queries taking less than a tenth of a millisecond which is a speedup of more than four orders of magnitude over Dijkstra's algorithm.
The preprocessing takes only a few minutes and produces little more auxiliary data than the input graph.
CH has been used successfully in many real world applications and quite a few open-source implementations exist\footnote{
\url{https://gist.github.com/PayasR/bc46af938195a827e42006c3f5544e4a}.
While this list is certainly not exhaustive and possibly not representative for what algorithms are used in cooperate contexts, it still paints a fairly clear picture on which techniques have reached a certain popularity in practice.
}.
CH is another fundamental algorithm for our work.
More specifically, we extend on PHAST~\cite{dgnw-phast-13} and RPHAST~\cite{delling_et_al:OASIcs:2011:3266}.
These algorithms reuse the CH preprocessing but modify the query phase.
PHAST supports \emph{one-to-all} queries from one source vertex to \emph{all} other vertices.
RPHAST is a variant for \emph{one-to-many} queries to a subset of vertices known in advance.
%
Another popular hierarchical speedup technique also used in practice~\cite{bingblog} is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02}.
It is based on a multi-level partition of the road network computed during preprocessing.
Queries are slightly slower than CH queries but still three to four orders of magnitude faster than Dijkstra's algorithm.
The memory consumption is even smaller than with CH.
Hierarchical speedup techniques have been combined with goal-directed search and A* in~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, these works also primarily focused on obtaining faster queries rather than simple and flexible approaches.

The speedup technique known to currently achieve the fastest queries is Hub Labels (HL)~\cite{DBLP:conf/esa/AbrahamDGW12,DBLP:conf/wea/DellingGW13}.
With HL, query times in less than a microsecond are possible.
This is only a couple of times the time of an uncached memory access.
This comes at the cost of an expensive preprocessing phase and a tremendous memory footprint, more than an order of magnitude larger than the input graph.
Despite the extremely fast query times, it appears that the trade-off offered by HL is, in practice, often not very attractive.
Simpler techniques like CH already offer queries fast enough to completely disappear behind other parts of practical applications (network latency, etc.).
Also, we are not aware of any works extending HL to extended problem models.

% extended application
\paragraph{Speedup Technique Adaptions for Extended Problems}
One of the earliest approaches to adapt speedup techniques to extended problems was based on A*.
In~\cite{dw-lbrdg-07}, ALT is used for routing on dynamic and time-dependent graphs.
The approach is conceptually similar to ours.
However, because the heuristic is ALT-based, the accuracy of the estimates is limited and the resulting query performance not competitive.
Further, the evaluation was performed with only synthetic traffic data.
With production-grade real-world traffic data, the problem becomes significantly harder~\cite{swz-sfert-21}.
Similar to the development for the classical problem, the approach was also combined with contraction~\cite{ndls-bastd-12,dn-crdtd-12}.
Even though this improves query running times, they obtained speedups are not competitive with purely hierarchical techniques.

CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of MLD~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
For this, a second, faster, preprocessing phase is introduced which takes at most a few seconds.
This phase is called the \emph{customization}.
It can be run regularly to update weights.
This enables the integration of live traffic and user preferences.
CRP was designed to support turn costs without any additional modifications.
Queries in CRP take at most a few milliseconds.
CRP is one of the few examples of a speedup technique designed for flexibility rather than maximum query performance.
However, its flexibility also has limits.
For example, integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
However, achieving reasonable memory consumption and query times was only possible by giving up exactness.
Further, TD-CRP can only compute approximate shortest distances rather than paths.

There is also lot of work that extends CH to more complex problem models.
In~\cite{gv-errnt-11}, turn information is integrated into CH.
The proposed approach achieves fast queries, but preprocessing becomes an order of magnitude slower compared to classical CH.
In~\cite{dsw-cch-15}, CH is extended to Customizable Contraction Hierarchies (CCH).
CCH also has a second preprocessing (customization) phase where weights can be altered.
This allows to support user preferences and live traffic with queries an order of magnitude faster than CRP.
Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
However, even with all improvements proposed in~\cite{bwzz-cchtc-20}, turn integration still causes a slowdown of at least factor three to the customization phase.
A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
Several papers~\cite{bdsv-tdch-09,bgns-tdcha-10,klsv-dtdch-10,bgsv-mtdtt-13} and an entire dissertation~\cite{b-tdrpc-14} have been published on the subject.
Different variants with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
Recently, a new approach has been published~\cite{swz-sfert-21} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.\footnote{In \url{https://github.com/kit-algo/rust_road_router}, the preprocessing alone has more than 10\,000 lines of code. In contrast, the classical CH can typically be realized within a few 100 lines of code.}

@@

Clearly, there is an enormous amount of efficient speedup techniques for routing in road networks.
Further, for many extended problems there are dedicated research results on extensions of certain techniques.
While many problems can be solved quite efficiently, we still believe that this situation is not satisfactory in practice.
For practical applications, a unified and extensible approach with manageable implementation complexity is often more important than the fastest query performance.
Therefore, we revisit the A* algorithm and build a flexible, unified framework for a multitude of routing problems.
The core of our approach is a new CH-based A* heuristic.
It allows for much tighter estimates than previous A* heuristics and thus significantly faster queries.
While the query running times of our approach are for the most part not competitive with approaches significantly tailored to specific problems, they are still sufficient for practical applications.
The remainder of this work is organized as follows:

We discuss notation and fundamental algorithms in Section~\ref{sec:preliminaries}.
In Section~\ref{sec:lazy-rphast}, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem.
Section~\ref{sec:astar_opts} contains several optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
Our main contribution, the CH-Potentials framework, is presented in Section~\ref{sec:framework}.
There, we first show how to use Lazy RPHAST to build a fast and tight heuristic for A* in road networks.
Based on this heuristic, we provide a unified framework (Section~\ref{sec:formal_framework} and~\ref{sec:chpot}) for a variety of practical route planning problems (Section~\ref{sec:applications}).
CH-Potentials can be applied to all of these problems without any modifications to the preprocessing.
In Section~\ref{sec:experiments}, we provide an extensive experimental evaluation analyzing the performance characteristics of our algorithms.
It shows that CH-Potentials achieve decent running times when tight lower bounds are available at preprocessing time.
However, the strength of our approach is not primarily in query running times but in its flexibility:
Problem extensions can be supported without complex algorithmic adjustments.
The price for each extension is a certain query slowdown depending on how tight the lower bounds used during preprocessing remain.

@@

The augmented graph has the property that for any two vertices $s$ and $t$, there always exists an \emph{up-down} $st$-path, i.e.\ a path  which first uses only edges from $\echu$ and then only edges from $\echd$, with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for an illustration.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Such an up-down path can be found by running Dijkstra's algorithm from $s$ on $\gchu$ and from $t$ on the reversed downward graph $\rgchd$ graph.
By construction, at least the highest ranked vertex on the up-down path must be in the intersection of the forward and backward search spaces.
Pseudocode for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
The set of vertices reachable in $\gchu$ and $\rgchd$ is called the \emph{CH search space} of a vertex.
The CH query performance strongly correlates with the size of this CH search space and the number of edges between the vertices in the search space.
Luckily, on road networks, the CH search space is small~\cite{gssv-erlrn-12,dgpw-crprn-13}.

@@

Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) is a variant of Contraction Hierarchies extended to a three-phase setup.
The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and the accelerated \emph{query} phase.
The preprocessing phase typically takes hours, the customization seconds, and the query fractions of a millisecond.
The preprocessing phase is independent of the weights and only uses the graph topology.
Weights are introduced in the customization phase.
Shortest paths are computed in the query phase.

The key idea for CCH is to use a vertex importance order that is independent of the weight function.
CCH use \emph{nested dissection} orders for this.
A small balanced vertex separator is computed and the separator vertices get the highest importance.
The remaining vertices are ordered by recursing on the independent cells obtained by removing the separator vertices.
This results in surprisingly good CH orders.
Regardless of the weight function, any shortest path between the cells has to use some of the separator vertices, therefore these vertices will always be ``important''.
Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19} in less than an hour even for continental-sized road networks.
Once the order was computed, the augmented graph $G^+$ is computed.
The standard CH preprocessing algorithms could be used for this but specialized, significantly faster CCH variants exist.
However, discussing them in detail is beyond the scope of this work.
See~\cite{dsw-cch-15,DBLP:journals/jea/Buchhold0W19} for details.
A CCH augmented graph fulfills all necessary properties for CH query algorithms.
The CH query algorithms can therefore be applied without modifications.
However, the CCH preprocessing grants some useful additional properties.

During preprocessing, an \emph{elimination tree} can be constructed.
The root of the elimination tree is the most important vertex.
For every other vertex, its parent is its least important upward neighbor.
In~\cite{bcrw-s-16}, it was proven that the set of ancestors of a vertex in the elimination tree is equal to its CH search space.
This makes it possible to explore the CCH search space with a more efficient shortest path algorithm:
For the forward search, start at the source and follow the elimination tree upward until the root is reached.
All outgoing edges of encountered vertices are relaxed.
The backward search works analogously but starts at the target.
Because this elimination tree-based algorithm does not require a queue, it is faster than a Dijkstra-based CCH query.
With this algorithm, CCH can consistently answer shortest path queries on continental sized road networks in fractions of a millisecond.
CH queries are still marginally faster for weight functions with a strong hierarchy such as travel times.
However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries turn out the faster variant.
Note that only nested dissection orderings admit an elimination tree of low height.
Thus, the elimination tree query can not be used with classical CH.

@@

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all vertices to one target vertex (or vice versa, the reverse case works analogously).
This is sometimes denoted as a \emph{all-to-one} (or \emph{one-to-all}) problem.
The preprocessing phase remains the same as for CH.
The query phase is split into two steps.
The first step is analogue to the CH query:
From $t$, all reachable vertices via reversed down-edges are explored as shown in Algorithm~\ref{algo:ch-backward}.
For the second step, PHAST utilizes an assignment of vertices into \emph{levels}.
These levels correspond to the importance ordering but allow multiple vertices in the same level.
However, no edge must connect two vertices within the same level.
Such a level assignment can be obtained by assigning all vertices without lower ranked neighbors to the lowest level.
All other vertices are then iteratively assigned to the next level above the highest level of any downward neighbor.

The main work of the second step consists in iterating over all CH levels from top to bottom.
In each iteration, all up-edges starting within the current level are relaxed in reverse.
After all levels are processed, the shortest distances from all vertices to $t$ were computed.
Pseudocode is provided in Algorithm~\ref{algo:phast}.
PHAST is faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures, better at utilizing data locality and, most importantly, can be parallelized very effectively.
% However, we will not consider parallelization in this paper.
See~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
% Using PHAST, we can also compute a tight A* heuristic.
% In the query phase, we first run PHAST to compute the distances from every vertex to $t$ with respect to $w$ and store the result in an array $H$.
% Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.

@@

We consider the combined running time of the target selection and each incremental query as the total running time to answer an incremental many-to-one query.
Since the target selection time is included, computing the distances to all vertices with Dijkstra or PHAST is too slow.
Also, since the source set is provided incrementally, RPHAST in its basic form is not well suited to our problem.
Fortunately, we can do better.
